---
title: "Writing unit tests for BaTFLED"
author: "Nathan Lazar"
date: "Aug 11, 2016"
output:
  pdf_document: default
  html_document:
    fig_caption: no
---
This tutorial is meant to walk you through writing unit tests using my package BaTFLED as an example.

## Trying out unit testing

Before we start playing with BaTFLED, lets first try writing some unit tests for simple functions. For this we'll need the package `testthat` written by R guru Hadley Wickem. 

```{r install package, message=F, warning=F, results='hide'}
pkgTest <- function(x) { # load packages and install if needed
  if (!require(x,character.only = TRUE)) {
    install.packages(x,dep=TRUE)
    if(!require(x,character.only = TRUE)) stop("Package not found")
}}
pkgTest('testthat')
```

Now define some simple function to play with.
(I stole most of this bit from R blogger John Myles White [here](https://www.r-bloggers.com/unit-testing-in-r-the-bare-minimum/))

```{r define function}
factorial <- function(n) {
  if(n==0) {return(1)} else {
    return(n * factorial(n-1)) 
  }
}
```

Some simple testing can be done with the functions `expect_equal`, `expect_true`, `expect_that`, etc. from the `testthat` package. The `test_that` function allows you to group tests and name them. The first argument is the name of the test which is displayed when a test fails.

```{r simple testing, eval=F}
expect_that(factorial(0), equals(1))
expect_true(factorial(3)==6)
  
test_that("factorial function works", {
  expect_equal(factorial(4), 24)
  expect_false(factorial(3)==5)
  expect_that(factorial(3)==5, is_false())
  expect_warning(log(-1))
})

test_that("this fails", {
  expect_equal(factorial(4), 12)
  expect_that(factorial(4), equals(12))
})

test_that("vector and matrix tests", {
  expect_equal(c(1,2,3), c(1,2,3))
  expect_equal(matrix(c(1,0,0,1), nrow=2, ncol=2), diag(1,2))
})
```

The real use case for this is to have a set of tests that you run on code as you're developing to make sure that you don't break anything. To do this, create a directory called `tests` in the BaTFLED3D directory that will store all of the test cases. In `tests` create a file `test1.R` that will contain your first set of tests. For example, you might have this:

```{r testing code, eval=F}
expect_that(2 ^ 2, equals(4))
expect_that(2 + 2 == 4, is_true())
expect_that(2 == 1, is_false())
expect_that(1, is_a('numeric'))
expect_that(print('Hello World!'), prints_text('Hello World!'))
expect_that(log('a'), throws_error())
expect_that(factorial(4), equals(24))
expect_false(factorial(3)==5)
expect_warning(log(-1))
expect_that(factorial(5), equals(110))
```

To run this set of tests, you can use `test_file` to just run this file or `test_dir` to run all the test files in the directory. For `test_dir`, the files containing tests have to be named `test*` and are executed in alpha-numeric order.

```{r run tests, eval=F} 
test_file('tests/test1.R')
test_dir('tests')
```

Also, you can keep watch on a directory and run tests anytime something is changed. First make a new directory called `code` and make a file containing your `factorial` function. Then, try running the code below and  changing or rewriting the factorial code to see if you've broken anything.

```{r watch directory, eval=F}
auto_test('code', 'tests')
```

## Writing tests for BaTFLED

Now lets try testing some real code in my package BaTFLED. First walk through the Rmarkdown document `BaTFLED_sim_CP.Rmd` to see that you can run the package. Then, try changing parameters for building the toy model or running BaTFLED and add some tests specific to BaTFLED3D to your `test1.R` file. Some examples are below, note that there are random aspects to BaTFLED, so to get a specific result for some functions, you may need to set the seed.

```{r writing BaTFLED tests, eval=F}
# Some unit tests of the input_data object
input.data <- input_data$new(mode1.X = matrix(rnorm(30), nrow=3, ncol=10),
                             mode2.X = matrix(rnorm(36), nrow=4, ncol=9), 
                             mode3.X = matrix(rnorm(40), nrow=5, ncol=8),
                             resp = array(rnorm(60), dim=c(3,4,5)))
expect_that(input.data$mode1.X, is_a('matrix'))
expect_equal(nrow(input.data$mode1.X), dim(input.data$resp)[[1]])

# Tests of get_data_params and mk_toy
cp.params <- get_data_params(list('decomp=CP'))
cp.toy <- mk_toy(data.params)
expect_that(cp.params$R > 0, is_true())
expect_that(is.null(cp.params$R1), is_true())
expect_that(cp.toy$core, is_a('array'))
expect_that(identical(cp.toy$resp, array(0,dim=c(cp.params$m1.rows, 
                                                 cp.params$m2.rows, 
                                                 cp.params$m3.rows))), 
            is_false())

# High level test to be run after training a model (not really a "unit" test):
# The lower bounds should always be increasing
if(model.params$lower.bnd)
  expect_that(sum((trained$lower.bnd[-1]-trained$lower.bnd[-reps])<0), equals(0))
```

Feel free to dig into the `BaTFLED3D/R/` folder, find a function or object and try writing a test for it. The documentation (`?<function or object>`) should have an explanation and some examples. Please tell me if you find anything difficult to understand, missing or not working. Some of the simpler objects/functions that could use tests are:

Objects:  
 - `input_data`  
 - `CP_model`  
 - `Tucker_model`  
 
Functions:  
 - `safe_log`  
 - `safe_prod`  
 - `exp_var`  
 - `mult_3d`: I plan on reimplementing this without the `rTensor` package so unit tests would be helpful  
 - `get_data_params`  
 - `get_model_params`  
 
Core functionality that has some random aspects:  
 - `train_Tucker`  
 - `train_CP`  
 - `update_mode[123]_CP`  
 - `update_mode[123]_Tucker`  
 - `update_core_Tucker`  
 
Some high level things that should be true after training a model:  
 - RMSEs should always be positive   
 - explained variation and correlations should be less than 1    